---
title: "HW03"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
First I will bring in my data and load my packages

```{r}
library(gapminder)
library(tidyverse)
library(ggplot2)
```

I'm going to focus on gdpPercap and get some more information like min and max per continent.
```{r min_max_gdpPercap}
gapminder %>%
  group_by(continent) %>% 
  summarize(min = min(gdpPercap),
            max = max(gdpPercap))
```

I can see that Asia has the highest gdpPercap and Africa has the lowest. 

Now I will look at the spread within each continent by looking at attributes like variance and standard deviation

```{r gdp_spread}
gapminder %>%
  group_by(continent) %>% 
  summarize(var = var(gdpPercap),
            sd = sd(gdpPercap))
```

Now I will calculate at trimmed mean for  lifeExp by year. I will trim the data removing 10% of the extreme values 

```{r trimmed_mean}
gapminder %>% 
  group_by(year) %>% 
  summarize(mean_lifeExp = mean(lifeExp, trim = .10))
```

Now I will look at how life expectancy is chaning over time between the continents

```{r change_in_LifeExp_over_time}
gapminder %>% 
  group_by(continent, year) 

```
Since this is difficult to visualize I will try a few graphs to make it easier to see.


```{r dot_Plot_LifeExp_Year_colorbycontinent}
gapminder %>% 
  ggplot(aes(year, lifeExp)) +
  geom_point(aes(color=continent))
```
This kind of works but lets see if something can work better.


```{r dotplot_LifeExp_year_separatebyContinent}
gapminder %>% 
  ggplot(aes(year,lifeExp)) +
  geom_point()+
  facet_wrap( ~ continent)
```
It is easier to visualize the trends this way. Maybe summarizing the data in a box plot will look even better.
```{r boxplot_LifeExp_Year_Continent}
gapminder %>% 
  ggplot(aes(year,lifeExp)) +
  geom_boxplot(aes(group=year))+
  facet_wrap( ~ continent)
```
I think this is the best plot because it shows a good summary of lifeExp for each continent over the year and is clear and easy to understand. 

Now I will look at low life expectancy by country over time. First, I will calculate the world wide median than count how many countries have a life expectancy lower than the median for each year.

```{r median_lifeExp_worldwide}
a <- gapminder %>% 
  summarize(median(lifeExp))
print(a)
 
  
```

The overall median is 60.7125. I will assign this value to variable a.
```{r life_expectancy_change_over_time}
below_median_life_expectancy <- gapminder %>% 
  group_by(country, year) %>% 
  filter(lifeExp < a) %>% select(c(country,year))

below_median_life_expectancy %>% ggplot(aes(x=year)) + geom_bar()
```
Now I will do some exploration to find a country with an intersting story. I think a country with a samll population could be interesting so I will try to find a country with the smallest average population
```{r min_population}
gapminder %>%
  group_by(country) %>% 
  mutate(min = min(pop)) %>%
  arrange(min)

```
I'm going to investigate Sao Tome and Principe. Instead of typing that out repeatedly I wil first assign it a nickname

```{r assign_variable}
STP <- "Sao Tome and Principe"
```

NOw I will look at the comparison between min and max pop
```{r miv_Vs_max_STP}
gapminder %>% 
  summarize(STP, min(pop),
            max(pop),
            var(pop))

```

That seems to be a very broad distribution. I will try to plot the data to see if there are any trends over time.

```{rlinegraph_popVstime}
gapminder %>%
  filter(country == STP) %>% 
  ggplot(aes(year, pop)) +
  geom_line()
  
```

It looks like the population steadily increased over the years. I will see if LifeExp follows the same trend

```{r linegraph_lifeExpVsTime}
gapminder %>%
  filter(country == STP) %>% 
  ggplot(aes(year, lifeExp)) +
  geom_line()
```

It does follow a somewhat similar trend in that it increases greatly over the years. 


I will try the extra exercises.
```{r}
gtbl <- tbl_df(gapminder)
ggplot(gtbl, aes(x = continent, y = gdpPercap, color = continent)) +
  geom_point() + (ggtitle("GDP per cap by continent"))
```

I have the plot but I am not sure why the table did not work
